RDA trait prediction tutorial

### 2023-02-08 rda_tutorial/RDAtraitPredictionTutorial.Rmd

---
title: "RDA trait prediction tutorial"
author: "KE Lotterhos"
date: "2023-01-20"
---

This tutorial accompanies the paper "The paradox of adaptive trait clines with non-clinal patterns in the underlying genes"

@@

Load the libraries and set your working directory:

```{r setup}

libraries_needed <- c("vegan", "LEA", "lfmm")

@@

#change code to your working directory
#knitr::opts_knit$set(root.dir = "~/Documents/GitHub/MVP-NonClinalAF/tutorial") 
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)

## Load the data

@@

## RDA trait prediction function

@@

### Understanding how the RDA is built on multiple regressions

@@

This visualization illustrates how there are unique ways in which environments are combined in the model to predict the pattern at each SNP.
```{r}
# look at the range of slopes
summary(as.numeric(Qr.coeff[,which(muts$causal)]))

@@

dev.new()
heatmap(t(Qr.coeff[,which(muts$causal)]),
        scale="none", col = cm.colors(length(brks)-1), breaks=brks)

>>>>

---
title: "RDA trait prediction tutorial"
author: "KE Lotterhos"
date: "2023-02-08"
output: pdf_document
---

This tutorial accompanies the paper "The paradox of adaptive trait clines with non-clinal patterns in the underlying genes" published in PNAS [add link].

# Abstract

Multivariate climate change presents an urgent need to understand how species adapt to complex environments. Population genetic theory predicts that loci under selection will form monotonic allele frequency clines with their selective environment, which has led to the wide use of genotype-environment associations (GEAs). However, the accuracy of GEA methods to identify adapted loci is limited, as shown in the main paper.

This tutorial shows how to apply a novel extension of multivariate ordination, which accurately predicted individual multivariate traits from genotype and environmental data on simulated data regardless of whether inference from GEAs was accurate.

# Install packages

@@

## Load the libraries:

```{r setup, eval=TRUE, echo=FALSE}

libraries_needed <- c("vegan", "LEA", "lfmm", "gplots")

@@

knitr::opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE) 
```

Don't forget to set your working directory!

# Background on the simulation

The six environmental variables are based on real data from western Canada and are shown below, clockwise from upper right: Clockwise from upper left: precipitation of driest month, precipitation of warmest quarter, mean annual temperature, precipitation of wettest month, mean temperature of wettest quarter, mean temperature of driest quarter. Each small square is a simulated individual, with its color representing its trait value in that environment.

![](multivar.png)

# Load the data

@@

# RDA trait prediction function

@@

# Understanding how the RDA is built on multiple regressions

@@

This visualization illustrates how there are unique ways in which environments are combined in the model to predict the pattern at each SNP.

```{r, fig.height=7, fig.width=8}
dim(Qr.coeff)
colnames(Qr.coeff) <- muts$mutname

# look at the range of coefficients in the multiple regression model
summary(as.numeric(Qr.coeff[,which(muts$causal)]))

@@

heatmap.2(t(Qr.coeff[,which(muts$causal)]),
        scale="none", 
        col = cm.colors(length(brks)-1), 
        breaks=brks,
        dendrogram = "column",
        Rowv=FALSE, #set this to "TRUE" if you would like to see which groups
        trace="none",
        key.title = "Coefficient in multiple\nregression model",
        ylab="SNPs",
        cexCol=1)
```

In the above heatmap, each row is a SNP. The SNPs are named according to their linkage group (1 through 10) and cumulative position in the genome (e.g. 9-448632 is on the 9th linkage group). Each linkage group was 50,000 bases long, so the cumulative position ranges from 1 to 500,000 over the 10 linkage groups. Each column in the heatmap is an environment, which is an explanatory variable in the model.

The color of the heatmap cells for a SNP shows the coefficients in the multiple regression model for each corresponding environment. In other words, the colors show how environments are combined in a multiple regression model to predict the patterns at that SNP on the landscape.

If we want to visualize clusters of SNPs that have similar coefficients in the multiple regression model, we can allow for clustering in the heatmap:

```{r, fig.height=7, fig.width=8}
heatmap.2(t(Qr.coeff[,which(muts$causal)]),
        scale="none", 
        col = cm.colors(length(brks)-1), 
        breaks=brks,
        dendrogram = "both",
        Rowv=TRUE, 
        trace="none",
        key.title = "Coefficient in multiple\nregression model",
        ylab="SNPs",
        cexCol=1)
```
